Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-27 src/tex/ms.tex
[showyourwork] overleaf sync

Finally, we publicly release all of our code to reproduce the results of this manuscript via GitHub\footnote{URL: {\url{https://github.com/TobiBu/sbi-chempy}}} and refer to Appendix \ref{sec:appendix_code_and_data} for a more extended overview of our code availability. All our datasets and network weights are publicly available on Zenodo.\footnote{URL: \url{https://zenodo.org/records/14883024}}

@@

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) from a given set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt} (which translates into a logarithmic error of %\variable{output/ape_NN_log.txt} dex)  
which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN})\footnote{Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14925307}{https://zenodo.org/records/14925307}.}. In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posterior estimator network.

@@

Note, the methods presented here can naturally be extended to any fast and flexible GCE model \citep[e.g.][and others]{Talbot1971,vice,omega2018}, not just \texttt{CHEMPY}.

@@

To facilitate a wider community's usage and contributions, we make use of the reproducibility software
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
\citep{Luger2021}, which leverages continuous integration to
programmatically download the data from
\href{https://zenodo.org/}{zenodo.org}, create the figures, and
compile the manuscript. Each figure caption contains two links: one
to the dataset stored on zenodo used in the corresponding figure,
and the other to the script used to make the figure (at the commit
corresponding to the current build of the manuscript). The git
repository associated to this study is publicly available at
\url{https://github.com/TobiBu/sbi-chempy}, and the release
v0.4.1 allows anyone to re-build the entire manuscript including rerunning all analysis. The datasets and neural network weights are stored at \url{https://doi.org/10.5281/zenodo.7343715} \textcolor{red}{change to final zenodo dataset!}.

>>>>

Finally, we publicly release all of our code to reproduce the results of this manuscript via GitHub\footnote{URL: {\url{https://github.com/TobiBu/sbi-chempy}}} and refer to Appendix \ref{sec:appendix_code_and_data} for a more extended overview of our code availability. All our datasets and network weights are publicly available on Zenodo.\footnote{URL: \url{https://zenodo.org/records/14925307}}

@@

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) from a given set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt} (which translates into a logarithmic error of \variable{output/ape_NN_log.txt} dex)  
which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN})\footnote{Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14925307}{https://zenodo.org/records/14925307}.}. In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posterior estimator network.

@@

Note, the methods presented here can naturally be extended to any fast and flexible GCE model \citep[e.g.][and others]{Talbot1971,vice,omega2018}, not just \texttt{CHEMPY} and can even be used to infer vastly different galactic parameters such as accretion events from abundance distributions of stars \citep[e.g.][]{Viterbo2024}.

@@

To facilitate a wider community's usage and contributions, we make use of the reproducibility software
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
\citep{Luger2021}, which leverages continuous integration to
programmatically download the data from
\href{https://zenodo.org/}{zenodo.org}, create the figures, and
compile the manuscript. Each figure caption contains two links: one
to the dataset stored on zenodo used in the corresponding figure,
and the other to the script used to make the figure (at the commit
corresponding to the current build of the manuscript). The git
repository associated to this study is publicly available at
\url{https://github.com/TobiBu/sbi-chempy}, and the release
v0.4.1 allows anyone to re-build the entire manuscript including rerunning all analysis. The datasets and neural network weights are stored at \url{https://zenodo.org/records/14925307}. The training and validation data can be found on zenodo as well under this link: \url{https://zenodo.org/records/14507221}.
